---
title: "Extracting data from the GREGoR data model"
output: html_notebook
---

This notebook has examples of how to extract data from workspace tables in the
[GREGoR data model](https://github.com/UW-GAC/gregor_data_models). We use the
[AnVIL R package](https://bioconductor.org/packages/AnVIL) to retrieve the 
data tables, and the [dplyr package](https://dplyr.tidyverse.org) to join tables 
in R. Both of these packages are installed in the AnVIL RStudio and Jupyter 
cloud environments.

# Identify a participant and family members

Let's say we are interested in retrieving data for a participant as well as all
family members. We can use the "participant" table to link all family members
by their family_id.

We read tables from a workspace with the `avtable` function.

```{r}
library(AnVIL)
library(dplyr)
workspace <- "GREGOR_COMBINED_CONSORTIUM_U07"
namespace <- "gregor-dcc"
participant <- avtable("participant", name=workspace, namespace=namespace)
```

Find the family_id for the participant, then filter the participant table to
all members of this family.

```{r}
my_participant_id <- "PMGRC-7-5-2"
my_family <- participant %>%
  filter(participant_id == my_participant_id) %>%
  select(family_id) %>%
  left_join(participant, by="family_id")
my_family
```

Find phenotypes for all family members.

```{r}
phenotype <- avtable("phenotype", name=workspace, namespace=namespace)
my_phenotype <- my_family %>%
  select(participant_id) %>%
  left_join(phenotype, by="participant_id")
my_phenotype
```

Use the family table to find more details about the family.

```{r}
family <- avtable("family", name=workspace, namespace=namespace)
my_family_details <- family %>%
  filter(family_id == unique(my_family$family_id))
my_family_details
```

# Find all analytes for participants in the family

The analyte table has information about the tissue that was used for the
experiments. We can filter this table by participant_id.

```{r}
analyte <- avtable("analyte", name=workspace, namespace=namespace)
my_analyte <- analyte %>%
  filter(participant_id %in% my_family$participant_id)
my_analyte
```


# Find experiments for all participants in the family

In the combined consortium workspace, the "experiment" table
includes all experiments across data types and links them to participant ids, 
making it more straightforward to find experiments for a given participant. We
still need to query the experiment tables for each data type represented to
find details on the experiments.

```{r}
experiment <- avtable("experiment", name=workspace, namespace=namespace)
my_experiment <- experiment %>%
  filter(participant_id %in% my_family$participant_id)
my_experiment

experiment_table_names <- unique(my_experiment$table_name)
experiment_tables <- lapply(experiment_table_names, avtable, name=workspace, namespace=namespace)
names(experiment_tables) <- experiment_table_names

experiment_details <- list()
for (t in experiment_table_names) {
  experiment_details[[t]] <- my_experiment %>%
    filter(table_name == t) %>%
    select(participant_id, id_in_table) %>%
    left_join(experiment_tables[[t]], by=c("id_in_table"=paste0(t, "_id")))
}
experiment_details
```



# Find all data files associated with a family

In the combined consortium workspace, the "aligned" table
includes all aligned read files across data types and links them to participant ids, 
making it more straightforward to find files for a given participant. We
still need to query the aligned tables for each data type represented to
find details on the alignment.

```{r}
aligned <- avtable("aligned", name=workspace, namespace=namespace)
my_aligned <- aligned %>%
  filter(participant_id %in% my_family$participant_id)
my_aligned

aligned_table_names <- unique(my_aligned$table_name)
aligned_tables <- lapply(aligned_table_names, avtable, name=workspace, namespace=namespace)
names(aligned_tables) <- aligned_table_names

aligned_details <- list()
for (t in aligned_table_names) {
  aligned_details[[t]] <- my_aligned %>%
    filter(table_name == t) %>%
    select(participant_id, id_in_table) %>%
    left_join(aligned_tables[[t]], by=c("id_in_table"=paste0(t, "_id")))
}
aligned_details
```
