---
title: "Joint callsets in the GREGoR data model"
output: html_notebook
---

This notebook has examples of how to identify samples in a joint callset in the
[GREGoR data model](https://github.com/UW-GAC/gregor_data_models). We use the
[AnVIL R package](https://bioconductor.org/packages/AnVIL) to retrieve the 
data tables, and the [dplyr package](https://dplyr.tidyverse.org) to join tables 
in R. Both of these packages are installed in the AnVIL RStudio and Jupyter 
cloud environments. We install the 
[AnvilDataModels](https://github.com/UW-GAC/AnvilDataModels) package from GitHub.

We read tables from a workspace with the `avtable` function.

```{r}
BiocManager::install("AnVILGCP")
remotes::install_github("UW-GAC/AnvilDataModels")
library(AnVIL)
library(dplyr)
library(AnvilDataModels)
workspace <- "GREGOR_COMBINED_CONSORTIUM_U11"
namespace <- "gregor-dcc"
set_table <- avtable("aligned_dna_short_read_set", name=workspace, namespace=namespace)
aligned_table <- avtable("aligned_dna_short_read", name=workspace, namespace=namespace)
experiment_table <- avtable("experiment_dna_short_read", name=workspace, namespace=namespace)
```

Find the set with a certain id, then use the `unnest_set_table` function from
the AnvilDataModels package to get the set table in long form (one row per id).

```{r}
joint_callset_id <- "GREGoR_DCC_A1_S3"
joint_callset <- set_table %>%
  filter(aligned_dna_short_read_set_id == joint_callset_id) %>%
  unnest_set_table()
```

The sample IDs in the VCF header are the experiment_sample_ids, which are in
the experiment table.

```{r}
sample_ids <- joint_callset %>%
  left_join(aligned_table, by="aligned_dna_short_read_id") %>%
  left_join(experiment_table, by="experiment_dna_short_read_id") %>%
  select(experiment_dna_short_read_id, experiment_sample_id)
```

We can also get the participant ids from the analyte table.

```{r}
analyte_table <- avtable("analyte", name=workspace, namespace=namespace)
participant_ids <- joint_callset %>%
  left_join(aligned_table, by="aligned_dna_short_read_id") %>%
  left_join(experiment_table, by="experiment_dna_short_read_id") %>%
  left_join(analyte_table, by="analyte_id") %>%
  select(participant_id, analyte_id, experiment_dna_short_read_id, experiment_sample_id)
```

Finally, find all the VCF files associated with the callset 
from the called_variants table.

```{r}
called_variants_table <- avtable("called_variants_dna_short_read", name=workspace, namespace=namespace)
vcf_files <- called_variants_table %>%
  filter(aligned_dna_short_read_set_id == joint_callset_id)
```
